HO unification from object language to meta language

### 2024-02-23 paper/main.tex
some typos

\noindent Under this context of instances a type-class solver is able to prove
the the following statement automatically by back-chaining.

@@

\noindent
Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
the concrete syntax to abstract, at the meta level, an expression
\elpiIn{e} over a variable \elpiIn{x}
is \elpiIn{x\ e}, and square brackets denote a list of
terms separated by comma. As an example we show the encoding of the Coq term
``$\forall y:\mathrm{t}.\mathrm{nfact}~ y~ 3$'':

@@

\noindent
We now illustrate the encoding of the three instances above as higher order
logic programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises;
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
we fail because of this ``higher order'' unification problem (for DTT)
is phrased as a first order unification problem in the meta language:
the two lists of terms have different lengths!

@@

\noindent
Since \elpiIn{Pm} is an higher order unification variable with \elpiIn{x}
in its scope, the unification problem (pa) admits one solution:

@@

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to definition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher order problems
restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
We call this unification procedure \Uo{}.

The equational theory of the meta language Elpi that we want to use to
implement a form of proof automation is strikingly similar, since it
it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves higher order problems in
$\mathcal{L}_\lambda$.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedure deal with are different. For example

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$ and we assume that the unification of our meta language is correct:

$$
t_1 \Ue t_2 \mapsto \sigma \Rightarrow \sigma t_1 \Ee \sigma t_2
$$

@@

% \printbibliography

>>>>

\noindent Under this context of instances a type-class solver is able to prove
the following statement automatically by back-chaining.

@@

\noindent
Following standard $\lambda$Prolog~\cite{miller_nadathur_2012}
the concrete syntax to abstract, at the meta level, an expression
\elpiIn{e} over a variable \elpiIn{x}
is <<\elpiIn{x\ e}>>, and square brackets denote a list of
terms separated by comma. As an example we show the encoding of the Coq term
<<\coqIn{~$\forall$~y:t, nfact y 3}>>:

@@

\noindent
We now illustrate the encoding of the three instances above as higher-order
logic-programming rules: capital letters denote rule
parameters; \elpiIn{:-} separates the rule's head from the premises;
\elpiIn{pi w\ p} introduces a fresh nominal constant \elpiIn{w}
for the premise \elpiIn{p}.

@@

\noindent
we fail because this ``higher order'' unification problem (for DTT)
is phrased as a first-order unification problem in the meta language:
the two lists of terms have different lengths!

@@

\noindent
Since \elpiIn{Pm} is an higher-order unification variable with \elpiIn{x}
in its scope, the unification problem (pa) admits one solution:

@@

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to definition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment $\mathcal{L}_\lambda$~\cite{miller92jsc}.
We call this unification procedure \Uo{}.

The equational theory of the meta language Elpi that we want to use to
implement a form of proof automation is strikingly similar, since it
it comprises $\eta\beta$ (for the meta language functions), and the
unification procedure \Ue{} solves higher-order problems in
$\mathcal{L}_\lambda$.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example

@@

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify with substitution
$\sigma$, we write $\sigma t$ for the application of the substitution to
$t$ and we assume that the unification of our meta language is correct:
%
$$
t_1 \Ue t_2 \mapsto \sigma \Rightarrow \sigma t_1 \Ee \sigma t_2
$$

@@

\printbibliography
